---
title: "UFO Data Cleaning"
output: 
  html_document:
    code_folding: hide
    
---

```{r, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

In this file we will show an example of data cleaning in R. You can click to see the code if you wish.

## Load our tools and data


First we need to load our tools - including the tidyverse and also read in the data, which we can do directly from a link. 

```{r}
library(tidyverse)
library(DT)
UFO <- read_csv(file = "https://docs.google.com/spreadsheets/d/e/2PACX-1vQp2WUFRZVTbmmJpvsP8CTNrk7EAMbSP9uPp_Doj6FMHAwyuZzasNrAEXcnf8dYBFElO-EHwiOmEsQd/pub?output=csv")
```

R allows us to check for any strange values, differences in column numbers etc. when we load the data. 
```{r}
problems(UFO)
```

Let's see if we see that in the Google Sheets version:

Row 30894: https://docs.google.com/spreadsheets/d/1T2ix2WGEsBXlsnowuu9GRyBVBgbIzid5CRsB4WQQyU4/edit?gid=49749752#gid=49749752&range=F30894

Row 39616: https://docs.google.com/spreadsheets/d/1T2ix2WGEsBXlsnowuu9GRyBVBgbIzid5CRsB4WQQyU4/edit?gid=49749752#gid=49749752&range=F39616

Row 65125:https://docs.google.com/spreadsheets/d/1T2ix2WGEsBXlsnowuu9GRyBVBgbIzid5CRsB4WQQyU4/edit?gid=49749752#gid=49749752&range=F65125

These would have been hard to find in Excel or Google Sheets!

## Examine the data

Take a minute to look at a birds eye view of the data. This will show us the top few rows.

```{r}
head(UFO)%>% datatable()
```

**Question**: We have how many variables? What are they?

<details> <summary> Answer </summary>

We have 11 variables. There is info about the date and time of a UFO siting, there is info about the location including the city, state, and country, latitude and longitude, as well as info about the shape of the sighting, the comments people wrote about them, and the date that the sighting was posted to this dataset.

</details>
<br>

R has many useful ways of viewing data. What makes it really special, is we can easily get summaries or subsets of the data that would be much more difficult in Google Sheets or Excel.


Say we want to see how many unique values there are for the country variable.

```{r}
count(UFO, country)
```

Nice, we can see that there are 5 different values and many rows that have `NA` for country.

You may have noticed in the data preview that the second row of data says it is from the state `tx` for Texas but the country info is missing.


If we look up "Lackland afb" - the value for the city variable - we see that it is indeed a real air force base in Texas in the United States. 


Let's try to save these rows where we have state data but we don't have country data.

We can use a list of state abbreviations that is already in R to do this! We will change the format so it matches the lowercase format in our data.

```{r}
lower_states <-tolower(state.abb)
lower_states
```


Let's take a look at more of these situations so we can verify that we want to keep the data:

```{r}
filter(UFO, state %in% lower_states & is.na(country)) %>% head() %>% datatable()
```



Now let's make a new country column based on the state column and if it matches the values in this list of state abbreviations `lower_states` and if it doesn't match we will have it keep the original country column value (which will be useful for values from Australia etc.)

```{r}
UFO <- UFO %>% mutate(
  country_new = case_when(state %in% lower_states ~ "US",
                                       .default = country
                          ))
UFO <- UFO %>% relocate(country_new, .before = country)

head(UFO)
```

Now let's filter for the US data. First let's try for the old version of the data which was missing `us` for some rows.

Let's see how many rows/sightings there are.

```{r}
filter(UFO, country == "us") %>% nrow() 
```


Now let's do it for the new country column.
Let's see how many rows/sightings there are.

```{r}
filter(UFO, country_new == "US") %>% nrow()
```


Wow, we saved a lot of United States sightings from being missed!

Now let's say we wanted only sightings from this 2014 in the US. This could be kinda tricky in Excel or Google Sheets. In R we can detect a pattern of "2014" and filter for it without needing to aadjust the `datetime` variable.

```{r}
US_2014 <-filter(UFO, str_detect(datetime, "2014") & country_new == "US")

US_2014
```

Now let's save what we have done, so we can use our cleaned data next time. We can save it as an R object so it is faster to import next time. Let's save all 3 objects that we created. 

```{r}
write_rds(UFO, file = "UFO_clean.rds")
write_rds(US_2014, file = "UFO_2014.rds")
write_rds(lower_states, file = "lower_states.rds")
```


